---
title: "NEFSC MSI Determinations"
subtitle: "Small Cetaceans and Pinnipeds"
date: today
publication-date: 2050-01-01
date-format: "MMMM DD, YYYY"
citeproc: true
bibliography: ../assets/MSI.yaml
csl: ../assets/apa.csl

author:
  - name: Lauren Noble
    email: lauren.noble@noaa.gov
    orcid: 0000-0001-111-1111
    affiliations:
      - ref: 1
  - name: Elizabeth Josephson
    email: elizabeth.josephson@noaa.gov
    affiliations:
      - ref: 1

affiliations:
  - id: 1
    name: Northesast Fisheries Science Center
    department: National Marine Fisheries Service
    address: 166 Water Street
    city: Woods Hole
    state: MA

execute: 
  echo: false
  warning: false
  message: false

format:
  html: 
    out-width: "90%"
    fig-width: 9
    fig-asp: 0.618
  nmfs-sar-pdf-typst:
    nmfs-region: World
    sar-year: 2025
    stock-name: "Stock B"
    common-name: "Pakicetus"
    genus-species: "Pakicetus sp."
    linenumbering: true
    keep-typ: false
    fig-width: 9
    fig-asp: 0.618
    fig-dpi: 600
---

```{r}
#| label: setup
#| include: false

library(dplyr)
library(ggplot2)
library(purrr)
library(fs)
library(gt)

# source all R scripts in the R/ folder
r_files <- fs::dir_ls(here::here("R"), glob = "*.R")
purrr::map(r_files, source)

# set the image device to ragg_png for better quality plots
knitr::opts_chunk$set(dev = "ragg_png")

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE
)

# make sure large numbers in text are formatted with commas
knitr::knit_hooks$set(
  inline = function(x) {
    if (!is.numeric(x)) {
      x
    } else {
      prettyNum(x, big.mark = ",")
    }
  }
)

options(scipen = 3, digits = 2)

# set the ggplot theme base font size and font family
theme_set(theme_minimal(base_size = 12, base_family = "Roboto"))

# function (for gt tables) to format numbers based on GAMMS specs
fmt_gamms_numbers <- function(x) {
  case_when(
    x <= 1 ~ formatC(x, format = "f", digits = 2),
    x > 1 & x <= 10 ~ formatC(x, format = "f", digits = 1),
    x > 10 ~ formatC(x, format = "f", digits = 0, big.mark = ","),
    TRUE ~ "format error"
  )
}
```

```{r}
#| label: initialize start year
year1<- 2019
```

## INTRODUCTION

The Marine Mammal Protection Act (MMPA) requires the National Marine Fisheries Service (NMFS) to estimate annual levels of human-caused serious injury and mortality to marine mammal stocks [@zotero-item-14418] and to categorize commercial fisheries based on their frequency of incidental serious injury and mortality of marine mammals [@zotero-item-14419]. Criteria for serious injury (SI) determinations were addressed at NMFS-convened workshops in 1997 and 2007 [@anglissDifferentiatingSeriousNonserious1998; @fisheriesDifferentiatingSeriousNonSerious2019], and in January 2012, the agency published national guidelines for distinguishing serious from non-serious injuries (NSI) of marine mammals. The procedural directive was again updated in February 2023 [@ProcessDistinguishingSerious2023]. According to this procedural directive, a serious injury (SI) is defined as an injury that has a \>50% chance of resulting in a mortality [@ProcessDistinguishingSerious2023]. A major goal of the guidelines is to establish national consistency and transparency in SI determinations. To implement the new guidelines, Science Center SI determination staff (SID) are required to annually review the observer (OBS) and at-sea monitor (ASM) records on all incidentally caught marine mammals that were released alive. This year, the Northeast Fisheries Science Center (NEFSC) also reviewed live stranding reports of small cetaceans and pinnipeds for serious injury determinations. Determinations made on these human interactions are independently reviewed by another center (Southwest Fisheries Science Center (SWFSC)), the Greater Atlantic Regional Fisheries Office (GARFO), and the Atlantic Scientific Review Group (ASRG) before final determinations are published in this document.

## METHODS

Electronic records of all small cetacean and pinniped bycatch that were coded as alive or condition unknown for the 5-year period (`r sprintf("%d", as.integer(year1))`-`r sprintf("%d", as.integer(year1+4))`) were extracted from the Northeast Fisheries Observer Program (NEFOP) database. A principal component of these records included OBS/ASM notes that provided information on entanglement characteristics (e.g., animal in cod end), crew handling (e.g., rope tied animal, animals lifted overboard), animal condition (e.g., cut on dorsal flank, some blood), and state of released animal (e.g., swam away quickly, swimming sluggishly at surface, immediately sank). These data were independently compared to small cetacean (S) and pinniped (P) criteria contained in the SI guidance document (Appendix) by 2 marine mammal researchers in the NEFSC Protected Species Division to determine their status (e.g., serious injury \[SI\], nonserious injury \[NSI\], uninjured \[UI\], Dead \[D\], or could not be determined \[CBD\]). The 2 evaluators compared their determinations, and all differences were discussed to obtain agreement. Details of each injury from the fishery observer programs and its determination are presented in Table 1. Final injury determinations and mortality events were tabulated to estimate the annual proportion of observed SI animals relative to the other observed determinations by gear type and species (Table 2). Proportions are used in stock assessment reports to partition the expanded bycatch estimate into serious injury vs. mortality (Hayes et al. 2024). Determinations for 2017-2021 have been previously published (Josephson and Lyssikatos 2023). Cases where a determination could not be made (CBD) were treated conservatively and included with the dead animals and hence represented in the final mortality and serious injury estimates reported in the annual stock assessment reports (Hayes et al. 2024). Offshore, migratory, and estuarine coastal bottlenose dolphin (*Tursiops truncatus truncatus*) injury determinations were analyzed and reported by the NOAA Fisheries Southeast Fisheries Science Center (Maze-Foley et al. 2019). Species codes and gear codes used in this report are contained, respectively, in Tables 3 and 4. The statistical area designations are presented in Figure 1. As a response to the ASRG and other agency reviews, the NMFS has started to include the serious injuries of marine mammals in stock assessment reports, as documented in the National Marine Mammal Health and Stranding database. In the Northeast, the majority of animals reported as live strandings with evidence of human interaction are pinnipeds. In the 2023 Marine Mammal Stock Assessment Report (2024), a table was provided in the gray seal *(Halichoerus grypus*) report that itemized gray seal serious injuries from 2017-2021 by type of interaction (fishery vs. other human interaction) and whether the animal was disentangled and released. Upon review of the 2022 data, and amid ongoing discussions with stranding network partners, managers, and scientific reviewers, it has been determined that a portion of the records that were in the database did not follow reporting guidelines that were in place at the time . Namely, these records represented sightings of entangled animals for which there was no stranding response. This resulted in unevenness in documentation of cases reported as entangled or injured animals and may also have resulted in double counting. We have made the decision, starting with the year 2022, to add the stranding database records to the serious injury determination process but only for cases that received stranding responses to be consistent with the national reporting guidelines. Table 5 itemizes and summarizes the 2022 stranding data that was reviewed, and Table 6 presents the 2017-2021 data previously reported in the gray seal stock assessment report. NOAA Fisheries is currently updating and clarifying the guidelines and forms for reporting live entangled pinnipeds, which will improve the data used in the future for serious injury determinations.

## RESULTS AND DISCUSSION

### Small Cetaceans

#### Observer Records

In 2019, 9 small cetacean records were reviewed. Seven of these were common dolphins (*Delphinus delphis*) with bottom trawl gear interactions, of which 4 were on the same trip. Upon review of observer descriptions and photos, 2 of the common dolphins were classified as dead and 2 as SI, with the remaining 3 meriting CBD status due to lack of information collected. The remaining 2 cetaceans reviewed in 2019 were recorded as an unidentified whale and unidentified dolphin. The “whale” was an animal brought on deck in the catch pile, probably in the cod end, described as being approximately 8 ft. in length, but the observer had only a quick view before the animal was hoisted overboard with a rope around the tailstock. The animal sank out of view when the line was cut, so it was most likely dead, but the species remains unknown. Two other bycatch mortalities on the same trip were harbor porpoises (*Phocoena phocoena*). The unidentified dolphin was perhaps a Risso’s dolphin (*Grampius griseus*) or pilot whale (*Globicephala sp*.) that had its flukes entangled in the tripper rope. After a line was cut, the animal was able to swim away, gear-free, apparently unharmed. However, since details were scant and species unknown, this record was classified as CBD.

In 2020 and 2021, observer coverage was greatly reduced due to COVID-19 restrictions. There were no takes recorded in the observer database as alive or condition unknown in 2020. In 2021, we reviewed 1 common dolphin, 1 white-sided dolphin (*Lagenorhynchus acutus*), 1 unidentified pilot whale and 1 unidentified dolphin for injury determination. The common dolphin and the white-sided dolphin were both determined to be SI as they were brought on deck. The pilot whale did not move while observed in the net but was cut from the net, and the captain of the vessel said it swam away. There was not enough information to make an injury determination. The unidentified dolphin, while not brought on board the vessel, fell onto the ramp when the net came out of the water, so it was also classified as SI under category S9 (body trauma).

In 2022, 3 common dolphins were recorded as alive or condition unknown after interactions with bottom trawl gear. From the observer’s description, 1 was clearly a fishery mortality. The other 2 were both brought on board with the catch before release, thus qualifying them as at least serious injuries. From the observer’s description, one of those animals could have been dead. Similarly, a harbor porpoise caught in a bottom trawl and brought on board before being released was determined to be SI. An unidentified pilot whale was caught in a bottom trawl but seen swimming freely in the net with the catch and released without being brought on board. The animal was observed rejoining its group and breathing and swimming normally. The animal was determined to be either UI or NSI.

*UPDATE - ADD 2023*

In summary, xx fishery bycatch records of small cetaceans from (`r sprintf("%d", as.integer(year1))`-`r sprintf("%d", as.integer(year1+4))`) were reviewed for injury determination. Of these, *UPDATE 8 were designated as SI, 1 as UI/NSI, 5 as CBD, and 15 as dead. Three common dolphins, 1 white-sided dolphin, 1 unidentified dolphin, and 2 Risso’s dolphins were seriously injured as a result of interacting with bottom trawl gear. One dolphin of unknown species, 1 common dolphin, 1 white-sided dolphin, and 1 harbor porpoise sustained serious injuries in sink gillnet gear. One unidentified marine mammal was classified as CBD in its interaction with trawl gear.*

#### *Stranding Records*

### Pinnipeds

#### Observer Records

#### Stranding Records

## ACKNOWLEDGEMENTS

We thank the onboard observers, at-sea monitors, and stranding responders for collecting the data and the staff in the Northeast Fisheries Science Center, Fisheries Monitoring Operations Branch, for their assistance in obtaining electronic copies of data logs and pictures required for our SI determinations.

## TABLES AND FIGURES

## REFERENCES CITED

## APPENDIX. TABLES 2 and 3 FROM

```{r}
#| label: insert table 1

tab1<-tibble(read.csv("../data/AppendixTable2.csv"))
tab1|>
  gt() |>
   tab_header(
    title = md("**Table 3**: Summary of Pinniped Injury Categories and Criteria")
     ) |>
     cols_align(
    align = "center",
    columns = c("Category", "Injury.Determination")
    ) |>
    cols_label(
    Injury.Information = "Injury/Information",
    Injury.Determination = "Injury Determination",
    Additional = "Additional factors for evaluating whether “case specific” injuries are serious or non-serious (additional factors at end of table)"
    )|>
  opt_table_font(size= 11
)|>
    tab_footnote(
    footnote = "This table includes on only those criteria determined to be serious injuries or case specific based on expert opinion at the 2007 Workshop (Andersen et al., 2008) and by pinniped experts on the NMFS Determination Staff working group. For the purposes of streamlining the information for the reader, criteria determined to be non-serious injuries are not included in this table.",
    locations = cells_body(columns = 3, rows = 1)
  )
```